---
title: "Econometrics case"
author: "Group 18"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
show_interpretation <- TRUE
```

# STEP 1 - Baseline Empirical Specification and Hypotheses
```{r echo=FALSE, results='asis'}
if (show_interpretation) {
  cat("We aim to estimate the total causal effect of RoL on income.

Latitude, continent dummies, and colonial origin are included as controls since they likely influence both institutions and income on their own, leaving them out would induce bias in our estimate of the RoL effect.

We decide to exclude Human Capital, as we believe it is a mediator. When you control for a mediator, you block the indirect causal channel and only estimate the direct effect. By excluding it we can estimate the total effect (direct + indirect) of RoL on income.

Our baseline model is as follows:")
}
```

$$\text{logY}_i = \beta_0 + \beta_1 \text{RoL}_i + \beta_2 \text{LAT}_i + \beta_3 D_{1i} + \beta_4 D_{2i} + \beta_5 D_{3i} + \beta_6 \text{FoB}_i + \beta_7 \text{FoF}_i + \mu_i$$

\textcolor{red}{Are we allowed to include our baseline model in our R output?}

```{r echo=FALSE, results='asis'}
  cat("Our assumptions are:\n
      - We know and have included all the confounders (no omitted variables)\n
      - No reverse causality\n
      - Human capital is a mediator and is not included\n
      ")
```

## Hypotheses

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.5} % Adds vertical padding for readability
\begin{tabular}{|p{2.5cm}|c|p{4cm}|l|p{3.5cm}|}
\hline
\textbf{Variable} & \textbf{Direction} & \textbf{Rationale} & \textbf{One/two-sided} & \textbf{$H_0, H_1$} \\ \hline
RoL & + & Better institutions lead to more income & One-sided & $H_0: \beta_1 \le 0$ \newline $H_1: \beta_1 > 0$ \\ \hline
LAT (latitude) & ? & Differences in geographical conditions & Two-sided & $H_0: \beta_2 = 0$ \newline $H_1: \beta_2 \neq 0$ \\ \hline
D1/D2/D3 (continent) & ? & Differences in regional conditions & Two-sided & $H_0: \beta_3, \beta_4, \beta_5 = 0$ \newline $H_1: \text{At least one} \neq 0$ \\ \hline
FoF, FoB (colonizer) & ? & Differences in former colonizer & Two-sided & $H_0: \beta_6, \beta_7 = 0$ \newline $H_1: \text{At least one} \neq 0$ \\ \hline
\end{tabular}
\end{table}

# STEP 2 - Load and Explore the Data
```{r}
# Load the data
data <- read.csv("Data.csv", sep = ';', row.names = 'X')
```

## STEP 2.1 - Overview of all the data
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Load the necessary libraries
library("psych")
library("dplyr")
library(corrplot)
library(ggplot2)
```
```{r echo=TRUE}
#Overview of all the data
head(data)
summary(data)
describe(data)
```
```{r echo=FALSE}
 #Compute covariance and correlation matrix (excluding region dummies)
data_corr <- data %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr, use = "complete.obs")
cor_matrix <- cor(data_corr, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr, use = "complete.obs")
cor(data_corr, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```


## STEP 2.2 - Overview for each region

```{r echo=FALSE}

attach(data)
index_africa <- D1 == TRUE
index_asia <- D2 == TRUE
index_america <- D3 == TRUE
data_africa <- data[index_africa, ]
data_asia <- data[index_asia, ]
data_america <- data[index_america, ]
```

### 1. Africa

#### 1.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_africa)
summary(data_africa)
describe(data_africa)
```

#### 1.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_af <- data_africa %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_af, use = "complete.obs")
cor_matrix <- cor(data_corr_af, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_af, use = "complete.obs")
cor(data_corr_af, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

### 2. Asia

#### 2.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_asia)
summary(data_asia)
describe(data_asia)
```

#### 2.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_as <- data_asia %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_as, use = "complete.obs")
cor_matrix <- cor(data_corr_as, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_as, use = "complete.obs")
cor(data_corr_as, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

### 3. America

#### 3.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_america)
summary(data_america)
describe(data_america)
```

#### 3.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_am <- data_america %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_am, use = "complete.obs")
cor_matrix <- cor(data_corr_am, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_am, use = "complete.obs")
cor(data_corr_am, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

## STEP 2.3 - Scatterplot of logY and RoL

```{r echo=FALSE}
ggplot(data = data, aes(x = RoL, y = logY)) +
  geom_point(color = "purple")
```
\newpage

# STEP 3 - Estimate and Analyze the Baseline Specification

## 3.1. Estimate the Baseline Specification with OLS

```{r echo=TRUE, warning=FALSE, message=FALSE}
# Fit the model
library(stargazer)
baseline <- lm(logY ~ RoL + LAT + D1 + D2 + D3 + FoF + FoB, data = data)
stargazer(data = baseline, type = "text", title = "OLS baseline")
```

## 3.2. Interpret the estimated coefficients

To test the significance of the coefficient of RoL we perform a one sided test. Our t-statistic is as follows:

$$t = \frac{0.821-0}{0.158} = 5.196$$
$$Critical\ value = t_{0.001; 54} = 3.232$$
$$5.196 > 3.232 \to We \ reject \ on \ the \ 0.1\% \ significance \ level$$

RoL: coefficient is significant on the 1% significance level. A one unit increase of RoL leads to a 82.1% increase in GDP per capita, ceteris paribus.

The controls are not significant on the 5% significance level. Since they are all two-sided tests, the t-statistics and p-values are calculated by the lm function.

Our regressors explain the variation in log GDP per capita for 65.5%. The included regressors are jointly significant on the 1% significance level. 

## 3.3 F-tests

```{r warning=FALSE, message=FALSE}
library(car)
f_test_region <- linearHypothesis(baseline, 
                                  c("D1 = 0", "D2 = 0", "D3 = 0"))
f_test_region
```

The continent dummies are jointly significant on the 1% significance level. At least one of the continent dummies has explanatory power.


```{r warning=FALSE, message=FALSE}
f_test_colonizer <- linearHypothesis(baseline, 
                                  c("FoF = 0", "FoB = 0"))
f_test_colonizer
```

The former colonizer dummies are not jointly significant on the 5% significance level. None of the former colonizer dummies have explanatory power.

```{r warning=FALSE, message=FALSE}
f_test_joint <- linearHypothesis(baseline, 
                                  c("LAT = 0", "D1 = 0", "D2 = 0", "D3 = 0", "FoF = 0", "FoB = 0"))
f_test_joint
```

The control variables are jointly significant. At least one of the control variables has explanatory power.

# STEP 4 - Evaluating Gauss-Markov Assumptions and Applying Remedial Measures

## 4.1. Nature of the regressors

Our dummy variables: D1, D2, D3, FoF and FoB are considered deterministic due to the fact that their values are limited to either 0 and 1 over repeated samples.\newline
LAT is a fixed coordinate, nothing can change that. This means LAT is deterministic.\newline
RoL is stochastic and endogenous. It almost certainly correlates with the error term. This is because log GDP per capita affects RoL (see dashed line in causal diagram).\newline
This implies that OLS becomes biased and inconsistent which means that the coefficient does not measure the true causal effect.
Remediation is required. This will be covered in step 4(g).

## 4.2. Distribution of Error Terms

### 4.2.1. Histogram of Residuals

```{r warning=FALSE}
library(ggplot2)
library(car)

# Extract residuals
residuals_baseline <- residuals(baseline)

# Histogram
ggplot(data.frame(residuals_baseline), aes(x = residuals_baseline)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "pink", color = "#FF2F92") +
  geom_density(color = "purple") +
  ggtitle("Histogram of Residuals")
```

### 4.2.2. Jarque-Bera Test for Normality

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tseries)
jarque.bera.test(residuals_baseline)
```

We cannot reject the H0 hypothesis on the 5% significance level. This means that we can not say that the error terms are not normality distributed on the 5% significance level.\newline
Side note: We have a small sample size. In small samples, the test may have low power, meaning it may fail to detect deviations from normality.\newline\newline
Normality does not have an effect on OLS point estimation. If the error term is not normally distributed, our t- and F-tests are unreliable because they assume normality of the error term. However, since we cannot say that the error term is not normally distributed, our t- and F-test should remain reliable.\newline
Remediation is not necessary in this step.

## 4.3. Multicollinearity

We compute the variance inflation factor:

```{r echo=FALSE}
library(car)
vif(baseline)
```
We see that we get high values for D1 and D3. We think this is because there are barely any countries that belong to the fourth category (D1 = 0, D2 = 0, D3 = 0). If we give each category a dummy variable, we get perfect multicollinearity. This situation doesn't differ much from the situation of perfect multicollinearity. This is why we think D1 and D3 have such high VIF's.\newline
Multicollinnearity is not a theoretical issue. It does not affect unbiasedness or consistency and remains the most efficient estimator. However, a high degree of multicollinearity can lead to larger variances of the estimators and thus wider confidence intervals and lower t-statistics. This decreases the power of inferential tests. Multicollinearity only leads to practical problems.\newline
Action should not be taken because multicollinearity mostly reflects a lack of information in the data.