---
title: "Econometrics case"
author: "Group 9 & Group 18"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
show_interpretation <- TRUE
```

# STEP 1 - Baseline Empirical Specification and Hypotheses
```{r echo=FALSE, results='asis'}
if (show_interpretation) {
  cat("We aim to estimate the total causal effect of RoL on income.

Latitude, continent dummies, and colonial origin are included as controls since they likely influence both institutions and income on their own, leaving them out would induce bias in our estimate of the RoL effect.

We decide to exclude Human Capital, as we believe it is a mediator. When you control for a mediator, you block the indirect causal channel and only estimate the direct effect. By excluding it we can estimate the total effect (direct + indirect) of RoL on income.

Our baseline model is as follows:")
}
```

$$\text{logY}_i = \beta_0 + \beta_1 \text{RoL}_i + \beta_2 \text{LAT}_i + \beta_3 D_{1i} + \beta_4 D_{2i} + \beta_5 D_{3i} + \beta_6 \text{FoB}_i + \beta_7 \text{FoF}_i + \mu_i$$


```{r echo=FALSE, results='asis'}
  cat("Our assumptions are:\n
      - We know and have included all the confounders (no omitted variables)\n
      - No reverse causality\n
      - Human capital is a mediator and is not included\n
      ")
```

## Hypotheses

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.5} % Adds vertical padding for readability
\begin{tabular}{|p{2.5cm}|c|p{4cm}|l|p{3.5cm}|}
\hline
\textbf{Variable} & \textbf{Direction} & \textbf{Rationale} & \textbf{One/two-sided} & \textbf{$H_0, H_1$} \\ \hline
RoL & + & Better institutions lead to more income & One-sided & $H_0: \beta_1 \le 0$ \newline $H_1: \beta_1 > 0$ \\ \hline
LAT (latitude) & ? & Differences in geographical conditions & Two-sided & $H_0: \beta_2 = 0$ \newline $H_1: \beta_2 \neq 0$ \\ \hline
D1/D2/D3 (continent) & ? & Differences in regional conditions & Two-sided & $H_0: \beta_3, \beta_4, \beta_5 = 0$ \newline $H_1: \text{At least one} \neq 0$ \\ \hline
FoF, FoB (colonizer) & ? & Differences in former colonizer & Two-sided & $H_0: \beta_6, \beta_7 = 0$ \newline $H_1: \text{At least one} \neq 0$ \\ \hline
\end{tabular}
\end{table}

# STEP 2 - Load and Explore the Data
```{r}
# Load the data
data <- read.csv("Data.csv", sep = ';', row.names = 'X')
```

## STEP 2.1 - Overview of all the data
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Load the necessary libraries
library("psych")
library("dplyr")
library(corrplot)
library(ggplot2)
```
```{r echo=TRUE}
#Overview of all the data
head(data)
summary(data)
describe(data)
```
```{r echo=FALSE}
 #Compute covariance and correlation matrix (excluding region dummies)
data_corr <- data %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr, use = "complete.obs")
cor_matrix <- cor(data_corr, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr, use = "complete.obs")
cor(data_corr, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```


## STEP 2.2 - Overview for each region

```{r echo=FALSE}

attach(data)
index_africa <- D1 == TRUE
index_asia <- D2 == TRUE
index_america <- D3 == TRUE
data_africa <- data[index_africa, ]
data_asia <- data[index_asia, ]
data_america <- data[index_america, ]
```

### 1. Africa

#### 1.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_africa)
summary(data_africa)
describe(data_africa)
```

#### 1.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_af <- data_africa %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_af, use = "complete.obs")
cor_matrix <- cor(data_corr_af, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_af, use = "complete.obs")
cor(data_corr_af, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

### 2. Asia

#### 2.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_asia)
summary(data_asia)
describe(data_asia)
```

#### 2.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_as <- data_asia %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_as, use = "complete.obs")
cor_matrix <- cor(data_corr_as, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_as, use = "complete.obs")
cor(data_corr_as, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

### 3. America

#### 3.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_america)
summary(data_america)
describe(data_america)
```

#### 3.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_am <- data_america %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_am, use = "complete.obs")
cor_matrix <- cor(data_corr_am, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_am, use = "complete.obs")
cor(data_corr_am, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

## STEP 2.3 - Scatterplot of logY and RoL

```{r echo=FALSE}
ggplot(data = data, aes(x = RoL, y = logY)) +
  geom_point(color = "purple")
```
\newpage

# STEP 3 - Estimate and Analyze the Baseline Specification

## 3.1. Estimate the Baseline Specification with OLS

```{r echo=TRUE, warning=FALSE, message=FALSE}
# Fit the model
library(stargazer)
baseline <- lm(logY ~ RoL + LAT + D1 + D2 + D3 + FoF + FoB, data = data)
stargazer(data = baseline, type = "text", title = "OLS baseline")
```

## 3.2. Interpret the estimated coefficients

To test the significance of the coefficient of RoL we perform a one sided test. Our t-statistic is as follows:

$$t = \frac{0.821-0}{0.158} = 5.196$$
$$Critical\ value = t_{0.001; 54} = 3.232$$
$$5.196 > 3.232 \to We \ reject \ on \ the \ 0.1\% \ significance \ level$$

RoL: coefficient is significant on the 1% significance level. A one unit increase of RoL leads to a 82.1% increase in GDP per capita, ceteris paribus. This is consistent with the institutional hypothesis.\newline

The controls are not significant on the 5% significance level. Since they are all two-sided tests, the t-statistics and p-values are calculated by the lm-function.\newline

Our regressors explain the variation in log GDP per capita for 65.5%. The included regressors are jointly significant on the 1% significance level. 

## 3.3 F-tests

```{r warning=FALSE, message=FALSE}
library(car)
f_test_region <- linearHypothesis(baseline, 
                                  c("D1 = 0", "D2 = 0", "D3 = 0"))
f_test_region
```

The continent dummies are jointly significant on the 1% significance level. At least one of the continent dummies has explanatory power.


```{r warning=FALSE, message=FALSE}
f_test_colonizer <- linearHypothesis(baseline, 
                                  c("FoF = 0", "FoB = 0"))
f_test_colonizer
```

The former colonizer dummies are not jointly significant on the 5% significance level. None of the former colonizer dummies have explanatory power.

```{r warning=FALSE, message=FALSE}
f_test_joint <- linearHypothesis(baseline, 
                                  c("LAT = 0", "D1 = 0", "D2 = 0", "D3 = 0", "FoF = 0", "FoB = 0"))
f_test_joint
```

The control variables are jointly significant. At least one of the control variables has explanatory power.

# STEP 4 - Evaluating Gauss-Markov Assumptions and Applying Remedial Measures

## 4.1. Nature of the regressors


Given that over repeated sampling, different countries can be observed, the values of D1, D2, D3, FoF, FoB and LAT are all stochastic. Furthermore, all of these have an effect on Human Capital, which is a mediator variable not included in the model and is therefore in the error term. This ultimately means that the aforementioned control variables are correlated with the error terms, and thus endogeneous.\newline
RoL is stochastic and endogenous. It almost certainly correlates with the error term. This is because log GDP per capita affects RoL (see dashed line in causal diagram).\newline
This implies that OLS becomes biased and inconsistent which means that the coefficients do not measure the true causal effect.
Remediation is required. This will be covered in step 4(g).

## 4.2. Distribution of Error Terms

### 4.2.1. Histogram of Residuals

```{r warning=FALSE}
library(ggplot2)
library(car)

# Extract residuals
residuals_baseline <- residuals(baseline)

# Histogram
ggplot(data.frame(residuals_baseline), aes(x = residuals_baseline)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "pink", color = "#FF2F92") +
  geom_density(color = "purple") +
  ggtitle("Histogram of Residuals")
```

### 4.2.2. Jarque-Bera Test for Normality

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tseries)
jarque.bera.test(residuals_baseline)
```

We cannot reject the H0 hypothesis on the 5% significance level. This means that we can not say that the error terms are not normality distributed on the 5% significance level.\newline
Side note: We have a small sample size. In small samples, the test may have low power, meaning it may fail to detect deviations from normality.\newline\newline
Normality does not have an effect on OLS point estimation. If the error term is not normally distributed, our t- and F-tests are unreliable because they assume normality of the error term. However, since we cannot say that the error term is not normally distributed, our t- and F-test should remain reliable.\newline
Remediation is not necessary in this step.

## 4.3. Multicollinearity

We compute the variance inflation factor:

```{r echo=FALSE}
library(car)
vif(baseline)
```
We see that we get high values for D1 and D3. We think this is because there are barely any countries that belong to the fourth category (D1 = 0, D2 = 0, D3 = 0). If we give each category a dummy variable, we get perfect multicollinearity. This situation doesn't differ much from the situation of perfect multicollinearity. This is why we think D1 and D3 have such high VIF's.\newline
Multicollinnearity is not a theoretical issue. It does not affect unbiasedness or consistency and remains the most efficient estimator. However, a high degree of multicollinearity can lead to larger variances of the estimators and thus wider confidence intervals and lower t-statistics. This decreases the power of inferential tests. Multicollinearity only leads to practical problems.\newline
Action should not be taken because multicollinearity mostly reflects a lack of information in the data.

\newpage

## 4.4 Heteroskedasticity
### Graphical detection of heteroskedasticity
#### Plot of Squared Residuals vs. Fitted Values

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.align='center'}
y_hat = baseline$fitted.values

res_sq <- residuals(baseline)^2

ggplot(data.frame(y_hat, res_sq), aes(x = y_hat, y = res_sq)) +
  geom_point(colour = "yellow", size = 1) +  # Points
  geom_smooth(method = "loess", se = FALSE, colour = "#FF2F92") +  # Trend line
  labs(
    title = "Squared Residuals vs. Fitted Values",
    x = "Fitted Values (log Y)",
    y = "Squared Residuals"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "grey90", colour = NA),   # Outer box
    panel.background = element_rect(fill = "purple", colour = NA),  # Plot area
    panel.grid.major = element_line(colour = "white"),              # Grid lines
    panel.grid.minor = element_blank()
  )
```

#### Plot of Squared Residuals vs. RoL

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.align='center'}
attach(data)

ggplot(data.frame(RoL, res_sq), aes(x = RoL, y = res_sq)) +
  geom_point(colour = "yellow", size = 1) +  # Points
  geom_smooth(method = "loess", se = FALSE, colour = "#FF2F92") +  # Trend line
  labs(
    title = "Squared Residuals vs. RoL",
    x = "RoL",
    y = "Squared Residuals"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "grey90", colour = NA),   # Outer box
    panel.background = element_rect(fill = "purple", colour = NA),  # Plot area
    panel.grid.major = element_line(colour = "white"),              # Grid lines
    panel.grid.minor = element_blank()
  )
```

#### Plot of Squared Residuals vs. LAT

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.align='center'}
attach(data)

ggplot(data.frame(LAT, res_sq), aes(x = LAT, y = res_sq)) +
  geom_point(colour = "yellow", size = 1) +  # Points
  geom_smooth(method = "loess", se = FALSE, colour = "#FF2F92") +  # Trend line
  labs(
    title = "Squared Residuals vs. LAT",
    x = "LAT",
    y = "Squared Residuals"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "grey90", colour = NA),   # Outer box
    panel.background = element_rect(fill = "purple", colour = NA),  # Plot area
    panel.grid.major = element_line(colour = "white"),              # Grid lines
    panel.grid.minor = element_blank()
  )
```

### Plot of Squared Residuals vs. Regions

```{r echo= FALSE, warning=FALSE, message=FALSE}
data <- data %>%
  mutate(Regions = case_when(
    D3 == 1 ~ 3,
    D2 == 1 ~ 2,
    D1 == 1 ~ 1,
    TRUE    ~ 0 
  ))
```

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.align='center'}
attach(data)

ggplot(data.frame(Regions, res_sq), aes(x = Regions, y = res_sq)) +
  geom_point(colour = "yellow", size = 1) +  # Points
  geom_smooth(method = "loess", se = FALSE, colour = "#FF2F92") +  # Trend line
  labs(
    title = "Squared Residuals vs. Regions",
    x = "Regions",
    y = "Squared Residuals"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "grey90", colour = NA),   # Outer box
    panel.background = element_rect(fill = "purple", colour = NA),  # Plot area
    panel.grid.major = element_line(colour = "white"),              # Grid lines
    panel.grid.minor = element_blank()
  )
```

On first sight, we notice a decrease in the spread of the residuals in the plots. For Africa (D1) we saw a slight increase contrary to the other regions. To conclude wether there is heteroskedadasticity, further testing is required. 

### Formal tests for heteroskedasticity

### Goldfeld-Quandt Test

Two-sided Goldfeld-Quandt test (hereinafter GQ1; no specified direction of variance change):
$$
H_0: \sigma_1^2 = \sigma_2^2 \quad \text{vs.} \quad H_1: \sigma_1^2 \neq \sigma_2^2
$$
One-sided GQ test for increasing variance (hereinafter GQ2):
$$
H_0: \sigma_2^2 \leq \sigma_1^2   \quad \text{vs.} \quad H_1: \sigma_2^2 > \sigma_1^2
$$
One-sided GQ test for decreasing variance (hereinafter GQ3):
$$
H_0: \sigma_2^2 \ge \sigma_1^2 \quad \text{vs.} \quad H_1: \sigma_2^2 < \sigma_1^2
$$

##### Goldfeld-Quandt Test by RoL

```{r warning=FALSE, message=FALSE, echo=FALSE}
library(lmtest)

GQ1_RoL=gqtest(baseline, fraction = 10, order.by = RoL, alternative = "two.sided")
GQ1_RoL

GQ2_RoL=gqtest(baseline, fraction = 10, order.by = RoL)
GQ2_RoL

GQ3_RoL=gqtest(baseline, fraction = 10, order.by = -RoL) # for decreasing variance
GQ3_RoL
```
Given that we saw a decreasing trend in variance, the third test is most relevant in our case. The p-value shows that we cannot reject the hypothesis of homoskedasticity. We assume that the residuals are normally distributed in order to perform this test. As shown in 4.2. this is the case.

#### White’s General Heteroskedasticity Test
```{r warning=FALSE, echo=FALSE}
# Auxiliary regression for White test (excluding interactions)
aux_model <- lm(res_sq ~ RoL + LAT + D1 + D2 + D3 + FoF + FoB + I(RoL^2) + I(LAT^2),
                data = data)

stargazer(aux_model, type="text")

# White's test statistic 
n <- nrow(data)
R2_aux <- summary(aux_model)$r.squared
white_stat <- n * R2_aux
df_white <- length(coef(aux_model)) - 1
p_value_white <- pchisq(white_stat, df_white, lower.tail = FALSE)

# Summary test result table 
data.frame(
  Statistic = c("Test Statistic", "Degrees of Freedom", "p-value"),
  Value = c(round(white_stat, 4), df_white, formatC(p_value_white, format = "e", digits = 4))
) |> knitr::kable(caption = "White's Heteroskedasticity Test Results")

summary(aux_model)$r.squared
```
The stargazer shows us that none of the coefficients are significant in the auxiliary regression. This leads us to believe that the error variance does not depend on the regressors. We dropped the interaction variables because of our small sample size. Including the interaction terms would reduce the degrees of freedom too much. This is important because the null hypothesis assumes an infinitely large sample size.

#### Brown-Forsythe Test by REGION
```{r echo=FALSE, warning=FALSE, message=FALSE}
# Group medians of residuals
attach(data)
med_resid_region <- tapply(residuals_baseline, Regions, median, na.rm = TRUE)

# Absolute deviations from group medians
z_region <- abs(residuals_baseline - med_resid_region[as.character(Regions)])

# One-way ANOVA on deviations
BF_region <- lm(z_region ~ Regions)

# Test H0 of equal variances across continents
anova_tab9 <- anova(BF_region)
F_stat3 <- anova_tab9["Regions", "F value"]
p_val3  <- anova_tab9["Regions", "Pr(>F)"]
df1_3   <- anova_tab9["Regions", "Df"]              # = G-1 = 3
df2_3   <- anova_tab9["Residuals", "Df"]           # = n-G = n-4

cat("Brown–Forsythe test (3 groups by region):\n")
cat("F =", round(F_stat3, 4), " df1 =", df1_3, " df2 =", df2_3,
    " p-value =", signif(p_val3, 4), "\n")

# Show group means of |deviations| for interpretation
tapply(z_region, Regions, mean, na.rm = TRUE)
```

#### Brown-Forsythe Test by COLONY
```{r echo=FALSE}
data <- data %>%
  mutate(Colony = case_when(
    FoF == 1 ~ 2,
    FoB == 1 ~ 1,
    TRUE    ~ 0 
  ))
```

```{r echo=FALSE}
# Group medians of residuals
attach(data)
med_resid_colony <- tapply(residuals_baseline, Colony, median, na.rm = TRUE)

# Absolute deviations from group medians
z_colony <- abs(residuals_baseline - med_resid_colony[as.character(Colony)])

# One-way ANOVA on deviations
BF_colony <- lm(z_colony ~ Colony)

# Test H0 of equal variances across continents
anova_tab4 <- anova(BF_colony)
F_stat4 <- anova_tab4["Colony", "F value"]
p_val4  <- anova_tab4["Colony", "Pr(>F)"]
df1_4   <- anova_tab4["Colony", "Df"]              # = G-1 = 3
df2_4   <- anova_tab4["Residuals", "Df"]           # = n-G = n-4

cat("Brown–Forsythe test (3 groups by colony):\n")
cat("F =", round(F_stat4, 4), " df1 =", df1_4, " df2 =", df2_4,
    " p-value =", signif(p_val4, 3), "\n")

# Show group means of |deviations| for interpretation
tapply(z_colony, Colony, mean, na.rm = TRUE)
```
Based on these tests we cannot reject the null hypothesis of homoskedasticity. Given that this is an F-test, the same remark as before about our sample size holds.

### Theoretical implications

Under heteroskedasticity OLS remains unbiased (and consistent), but becomes inefficient to GLS. Inference becomes invalid, because the variance formula used in OLS is no longer correct and needs to be adjusted.

### Addressing "pure" heteroskedasticity
Disclaimer: heteroskedasticity was not a problem. 

#### White correction: Heteroskedasticity-robust standard errors

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(sandwich)
baseline_White = baseline
White_se=sqrt(diag(vcovHC(baseline_White)))
stargazer(baseline, baseline_White,
          type = "text",
          style = "all",
          se = list(NULL, White_se),
          column.labels = c("Baseline", "White SE"))
```

#### EGLS (Feasible GLS) estimator
```{r echo=FALSE, warning=FALSE}
library(modelsummary)
library(sandwich)
library(broom)

# 1. Extract the fitted values
y_hat <- fitted.values(baseline)

# 2. Define weights: if variance ~ 1/y_hat, then weight = y_hat
w_final <- y_hat

# 3. Run the WLS model
EGLS_RoL <- lm(logY ~ RoL + LAT + D1 + D2 + D3 + FoB + FoF, 
                 data = data, 
                 weights = w_final)

# 5) Show results
stargazer(baseline, EGLS_RoL,
          type = "text",
          style = "all",
          column.labels = c("Baseline", "EGLS by RoL"))

```
We see that the variance structure is decreasing with an increasing logY. So we assume that the error variance is proportional to 1/logY. We multiplied each value with the square root of its fitted value. This way higher fitted values get a higher weight because they have a lower variance.
We see that with the EGLS model the R squared drops and the standards errors of the estimators dramatically increase. This may indicate that the given transformation was not relevant.

### Theoretically comparing OLS to EGLS

The efficiency and consistency of EGLS depends on the correctness of the assumed heteroskedasticity pattern. EGLS standard errors are not necessarily lower than OLS standard errors as OLS incorrectly assumes homoskedasticity. In the case of heteroskedasticity, EGLS is consistent and asymptotically efficient.

## 4.5. Autocorrelation

### Graphical inspection of Autocorrelation
```{r echo=FALSE}
# Create dataframe with residuals and RoL
res_RoL_df <- data.frame(RoL = RoL, Residuals = residuals(baseline))

# Order by RoL
res_RoL_df <- res_RoL_df[order(res_RoL_df$RoL), ]
res_RoL_df$Index <- 1:nrow(res_RoL_df)

# Plot
ggplot(res_RoL_df, aes(x = Index, y = Residuals)) +
  geom_point(colour = "yellow", size = 1) +  # Points
  geom_smooth(method = "loess", se = FALSE, colour = "#FF2F92") +  # Trend line
  labs(
    title = "Residuals vs. Observation Index (Ordered by RoL)",
    x = "Ordered Index by RoL",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "grey90", colour = NA),   # Outer box
    panel.background = element_rect(fill = "purple", colour = NA),  # Plot area
    panel.grid.major = element_line(colour = "white"),              # Grid lines
    panel.grid.minor = element_blank()
  )

```
Based on this graph, it is difficult to see if there is some sort of negative correlation or no autocorrelation at all. This needs to be formally tested.

### Formal autocorrelation tests 

### Runs test
```{r echo=FALSE, warning=FALSE, message=FALSE}
# Order data by EDUC
ordered_data <- data[order(data$RoL), ]
res_OR <- residuals(lm(logY ~ RoL + LAT + D1 + D2 + D3 + FoF + FoB, data = ordered_data))

# Create sign vector (+1 for positive, -1 for negative residuals)
signs <- ifelse(res_OR >= 0, 1, -1)

# Count runs
runs_count <- function(signs) {
  r <- 1
  for (i in 2:length(signs)) {
    if (signs[i] != signs[i - 1]) {
      r <- r + 1
    }
  }
  return(r)
}

R <- runs_count(signs)
N1 <- sum(signs == 1)
N2 <- sum(signs == -1)
N <- N1 + N2
E_R <- 2 * N1 * N2 / N + 1
s_R <- sqrt(2 * N1 * N2 * (2 * N1 * N2 - N) / (N^2 * (N - 1)))

results_R <- c(R, E_R, E_R - 1.96 * s_R, E_R + 1.96 * s_R)
names(results_R) <- c("Observed Runs", "Expected Runs", "95% Lower Bound", "95% Upper Bound")
results_R

# Comparison using runs.test:
library(tseries)
sign_residuals <- factor(signs)
runs_result <- runs.test(sign_residuals)
runs_result
```

#### Durbin-Watson d-statistic
```{r echo=FALSE}
baseline_OR = lm(logY ~ RoL + LAT + D1 + D2 + D3 + FoB + FoF, data = ordered_data)

library(lmtest)
dwtest(baseline_OR)
```

#### Breusch-Godfrey LM test
```{r echo=FALSE, message=FALSE, warning=FALSE}
bgtest(baseline_OR, order = 1)
```

These tests conclude that there is not autocorrelation.

### Specific assumption of the previous test

### Runs test

The runs test has no major assumptions, as it is non-parametric and thus does not have any functional form related to it.

### Durbin-Watson d-test

This test assumes non-stachastic estimators, which is not the case. Nevertheless, there are usually no big implication.
Furthermore, the error terms should follow a AR(1) process. If this not true, like in this case, its power to detect autocorrelation decreases.
Thridly, the error terms should be normally distributed, which is the case.

The most critical assumptions for the Durbin-Watson test is that the regression model does not include lagged dependent variables.

### Breusch-Godfrey LM test

The Breusch-Godfrey LM test assumes that the data follows an AR(p) pattern. It is less powerful in small samples.

### Interpretation and remediation

If autocorrelation would be present in cross-sectional data, this would typically implicate modelmisspecification.
Under autocorrelation, OLS estimators remain unbiased and normally distributed, but the standard variance formula is incorrect. This leads to biased standard errors and unreliable inference. In addition, OLS is no longer efficient, as GLS provides estimators with lower variance.
In case remediation is necessary, the more efficient (E)GLS model can remove autocorrelation, or in case of non-stationary data, first-differencing can be considered. In order to keep using OLS, the formula for the standard errors would have to be corrected using the Newey-West correction. However, as our sample size is quite small, doing nothing might be reasonable. Autocorrelation could also be a signal of misspecification, which we will discuss in the next part.

## 4.6. Model Specification

### Alternative specifications
#### Extend the model by allowing the effect of RoL to differ across continents

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Log-linear transformation of the baseline

baseline_inter_RoL <- lm(logY ~ RoL + LAT + D1 + D2 + D3 + FoF + FoB + RoL*D1 + RoL*D2 + RoL*D3 , data = data)

# Stargazer output
library(stargazer)
stargazer(baseline, baseline_inter_RoL,
          type = "text",
          title = "Baseline with RoL across continents",
          column.labels = c("Level", "Log-Level"),
          align = TRUE,
          no.space = TRUE)
```

```{r echo=FALSE}
resettest(baseline_inter_RoL, power = 2:3, type = "fitted")
```

```{r warning=FALSE, message=FALSE}
f_test_inter_RoL <- linearHypothesis(baseline_inter_RoL, 
                                  c("RoL:D1 = 0", "RoL:D2 = 0", "RoL:D3 = 0"))
f_test_inter_RoL
```

Since the p-value (0.1209) is greater than 0.05, we fail to reject the null hypothesis. The interaction terms are not jointly significant. This suggests that the impact of the Rule of Law on logY does not statistically differ significantly across the different regions defined by D1,D2, and D3.

#### Extend the specification by including human capital
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Fit the model with higher-order EDUC
baseline_HC <- lm(logY ~ RoL + LAT + D1 + D2 + D3 + FoF + FoB + HC, data = data)

# Use stargazer to output the results, omitting region dummies
library(stargazer)
stargazer(baseline_HC, type = "text", title = "OLS Baseline with Human capital", 
          align = TRUE)
```
```{r echo=FALSE}
resettest(baseline_HC, power = 2:3, type = "fitted")
```

```{r echo=FALSE}
# 1. Extract the summary coefficients table
stats <- summary(baseline)$coefficients

# 2. Pull the specific values for RoL
mag   <- stats["RoL", "Estimate"]
p_val <- stats["RoL", "Pr(>|t|)"]
sign  <- ifelse(mag > 0, "+", "-")

# 3. Print the results clearly
cat("RoL without HC \n")
cat("Magnitude:", round(mag, 4), "\n")
cat("Sign:     ", sign, "\n")
cat("p-value:  ", signif(p_val, 4), "\n \n")

# 1. Extract the summary coefficients table
stats_HC <- summary(baseline_HC)$coefficients

# 2. Pull the specific values for RoL
mag_HC   <- stats_HC["RoL", "Estimate"]
p_val_HC <- stats_HC["RoL", "Pr(>|t|)"]
sign_HC  <- ifelse(mag_HC > 0, "+", "-")

# 3. Print the results clearly
cat("RoL with HC \n")
cat("Magnitude RoL:", round(mag_HC, 4)      , "     Magnitude HC", stats_HC["HC", "Estimate"] , "\n")
cat("Sign RoL:     ", sign_HC               , "          Sign HC:       ", ifelse(stats_HC["HC", "Estimate"] > 0, "+", "-"), "\n")
cat("p-value RoL:  ", signif(p_val_HC, 4)   , "   p-value HC:  ", signif(stats_HC["HC", "Pr(>|t|)"], 4), "\n")
```
The coefficient for RoL dropped significantly (from 0.8207 to 0.4277) when HC was added. This suggests a strong relationship between institutions and human capital.

HC as a Confounder: If we view HC as a confounder, it means HC affects both the quality of institutions and GDP. In this view, the baseline model was overestimating the effect of RoL because it was "soaking up" the effect of omitted human capital.
HC as a Mediator: If we view HC as a mediator, it means good institutions (RoL) lead to better HC, which in turn leads to higher logY. In this case, the drop in the coefficient is expected because HC is the channel through which institutions work. By including it, we are measuring only the direct effect of RoL, ignoring its indirect effect via education. 

Conclusion:
The Baseline Model without Human Capital stands as the most appropriate specification. Econometrically, this choice is supported by the joint insignificance of the continent interaction terms (with an F-test p-value of 0.1209), these extra variables add unnecessary complexity without improving the model’s explanatory power. Furthermore, including Human Capital as a regressor introduces a "bad control" problem. Adding it to the model "soaks up" the influence of the Rule of Law, reducing its coefficient from 0.82 to 0.43 and obscuring the true magnitude of its impact.

Economically, institutions are recognized as "deep determinants" of growth that precede and incentivize investments in education and physical capital. The Baseline Model captures the total effect of institutions, which includes both their direct influence on productivity and their indirect influence through the development of a skilled workforce. By contrast, the model with Human Capital only measures the narrow direct effect, which underrepresents the actual role of the Rule of Law in economic development. Additionally, Human Capital is often endogenous—where higher income levels lead to better education—which can introduce simultaneity bias. Therefore, the Baseline Model remains the most robust and theoretically sound representation of how institutional quality drives economic performance.


## 4.7. Endogeneity
### 2SLS estimation
```{r}

# First stage estimation
first_stage=lm(logY ~ logSM + logPD + RoL + LAT + D1 + D2 + D3 + FoB + FoF, data = data)
summary(first_stage)

library(car)
linearHypothesis(first_stage, matchCoefs(first_stage, "logSM"))

# Second stage estimation
library(AER)
twosls=ivreg(WAGE ~ EDUC + AGE + RACE + MARRIED + URBAN + REGION_factor | QOB_factor + AGE + RACE + MARRIED + URBAN + REGION_factor, data = data)
summary (twosls, diagnostics=TRUE) #NIET MEER SIGNIFICANT --> WIJST OP WEAK INSTRUMENT | F is below 10 (fail on relevance condition)

stargazer(baseline, twosls,
          type = "text",
          title = "Baseline OLS vs 2SLS Estimates",
          column.labels = c("OLS", "2SLS"),
          align = TRUE,
          keep.stat = c("n", "rsq", "adj.rsq", "f"),
          no.space = TRUE)
```

### Testing for Endogeneity
```{r}
res_first=first_stage$residuals

haus=lm(WAGE ~ EDUC + AGE + RACE + MARRIED + URBAN + REGION_factor + res_first, data=data)
summary (haus)
```