---
title: "Econometrics case"
author: "Group 18"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
show_interpretation <- TRUE
```

# STEP 1 - Baseline Empirical Specification and Hypotheses
```{r echo=FALSE, results='asis'}
if (show_interpretation) {
  cat("We aim to estimate the total causal effect of RoL on income.

Latitude, continent dummies, and colonial origin are included as controls since they likely influence both institutions and income on their own, leaving them out would induce bias in our estimate of the RoL effect.

We decide to exclude Human Capital, as we believe it is a mediator. When you control for a mediator, you block the indirect causal channel and only estimate the direct effect. By excluding it we can estimate the total effect (direct + indirect) of RoL on income.

Our baseline model is as follows:")
}
```

$$\text{logY}_i = \beta_0 + \beta_1 \text{RoL}_i + \beta_2 \text{LAT}_i + \beta_3 D_{1i} + \beta_4 D_{2i} + \beta_5 D_{3i} + \beta_6 \text{FoB}_i + \beta_7 \text{FoF}_i + \mu_i$$

\textcolor{red}{Are we allowed to include our baseline model in our R output?}

```{r echo=FALSE, results='asis'}
  cat("Our assumptions are:\n
      - We know and have included all the confounders (no omitted variables)\n
      - No reverse causality\n
      - Human capital is a mediator and is not included\n
      ")
```

## Hypotheses

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.5} % Adds vertical padding for readability
\begin{tabular}{|p{2.5cm}|c|p{4cm}|l|p{3.5cm}|}
\hline
\textbf{Variable} & \textbf{Direction} & \textbf{Rationale} & \textbf{One/two-sided} & \textbf{$H_0, H_1$} \\ \hline
RoL & + & Better institutions lead to more income & One-sided & $H_0: \beta_1 \le 0$ \newline $H_1: \beta_1 > 0$ \\ \hline
LAT (latitude) & ? & Differences in geographical conditions & Two-sided & $H_0: \beta_2 = 0$ \newline $H_1: \beta_2 \neq 0$ \\ \hline
D1/D2/D3 (continent) & ? & Differences in regional conditions & Two-sided & $H_0: \beta_3, \beta_4, \beta_5 = 0$ \newline $H_1: \text{At least one} \neq 0$ \\ \hline
FoF, FoB (colonizer) & ? & Differences in former colonizer & Two-sided & $H_0: \beta_6, \beta_7 = 0$ \newline $H_1: \text{At least one} \neq 0$ \\ \hline
\end{tabular}
\end{table}

# STEP 2 - Load and Explore the Data
```{r}
# Load the data
data <- read.csv("Data.csv", sep = ';', row.names = 'X')
```

## STEP 2.1 - Overview of all the data
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Load the necessary libraries
library("psych")
library("dplyr")
library(corrplot)
```
```{r echo=TRUE}
#Overview of all the data
head(data)
summary(data)
describe(data)
```
```{r echo=FALSE}
 #Compute covariance and correlation matrix (excluding region dummies)
data_corr <- data %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr, use = "complete.obs")
cor_matrix <- cor(data_corr, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr, use = "complete.obs")
cor(data_corr, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```


## STEP 2.2 - Overview for each region

```{r echo=FALSE}

attach(data)
index_africa <- D1 == TRUE
index_asia <- D2 == TRUE
index_america <- D3 == TRUE
data_africa <- data[index_africa, ]
data_asia <- data[index_asia, ]
data_america <- data[index_america, ]
```

### 1. Africa

#### 1.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_africa)
summary(data_africa)
describe(data_africa)
```

#### 1.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_af <- data_africa %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_af, use = "complete.obs")
cor_matrix <- cor(data_corr_af, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_af, use = "complete.obs")
cor(data_corr_af, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

### 2. Asia

#### 2.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_asia)
summary(data_asia)
describe(data_asia)
```

#### 2.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_as <- data_asia %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_as, use = "complete.obs")
cor_matrix <- cor(data_corr_as, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_as, use = "complete.obs")
cor(data_corr_as, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

### 3. America

#### 3.1 Overview

```{r echo=TRUE}
#Overview of all the data
head(data_america)
summary(data_america)
describe(data_america)
```

#### 3.2 Covariance and correlation matrix

```{r echo=FALSE}
#Compute covariance and correlation matrix (excluding region dummies)
data_corr_am <- data_america %>%
  select(logY, RoL, LAT, FoF, FoB)

cov_matrix <- cov(data_corr_am, use = "complete.obs")
cor_matrix <- cor(data_corr_am, use = "complete.obs")
```
```{r echo=TRUE}
cov(data_corr_am, use = "complete.obs")
cor(data_corr_am, use = "complete.obs")
#Visualize the correlation matrix (excluding region dummies)
corrplot(cor_matrix, method = "color")
```

# STEP 3 - Estimate and Analyze the Baseline Specification

## 3.1. Estimate the Baseline Specification with OLS

```{r echo=TRUE}

```